The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-15 talk/vmil2012/paper.tex
more tweaking

All measurements presented in this section do not take garbage collection of resume data and machine code into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and data that is generated from the optimized traces.

@@

Figure~\ref{fig:benchmarks} extends Figure~\ref{fig:guard_percent} and summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The Figure shows that the overall optimization rate for
operations, which is between 69.4\% and 83.89\%, of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations, are very similar. This indicates that the optimizer can remove
most of the guards, but after the optimization pass these still account for
15.2\% to 20.2\% of the operations being compiled and later executed.
The frequency of guard operations makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
quickly.

@@

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the optimized traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
enough that a bridge is compiled for them, only 1.2\% to 3.6\% of all guards
fail often enough that a bridge is compiled. Also, of all failing guards a few fail extremely often
and most fail rarely. The results emphasize that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

@@

SPUR~\cite{bebenita_spur:_2010} is a tracing JIT compiler
for a C\# virtual machine.
It handles guards by always generating code for every one of them
that transfers control back to the unoptimized code.
Since the transfer code needs to reconstruct the stack frames
of the unoptimized code,
the transfer code is quite large.

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a guard failure using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions the plans to switch to compressed snapshots to further reduce
redundancy. The approach of not creating snapshots at all for every guard is
orthogonal to the resume data compression presented in this paper and could be
reused within RPython to improve the memory usage further.

Linking side exits to pieces of later compiled machine code was described first
in the context of Dynamo~\cite{Bala:2000wv} under the name of Fragment Linking.
Once a new hot trace is emitted into the fragment cache it is linked to side
exit that led to the compilation of the fragment. Fragment Linking avoids the
performance penalty involved in leaving the compiled code. Fragment linking
also allows to remove compensation code associated to the linked fragments that
would have been required to restored the execution state on the side exit.

Gal et. al~\cite{Gal:2006} describe how in the HotpathVM they experimented
with having one generic compensation code block, like the RPython JIT, that
uses a register variable mapping to restore the interpreter state. Later this
was replaced by generating compensation code for each guard which produced a
lower overhead in their benchmarks. HotpathVM also records secondary traces
starting from failing guards that are connected directly to the original trace.
Secondary traces are compiled by first restoring the register allocator state to
the state at the side exit. The information is retrieved from a mapping stored
in the guard that maps machine level registers and stack to Java level stack
and variables.

@@

\section{Conclusion}
\label{sec:Conclusion}
In this paper we have concentrated on guards, an operation found in
tracing just-in-time compilers and used to denote points of possible control
flow divergence in recorded traces.
Based on the observation that guards are a frequent operation in traces and
that they do not fail often, we described how they have been implemented in the
high and low level components of RPython's tracing JIT compiler.

Additionally we have presented experimental data collected using the standard PyPy
benchmark set to evaluate previous observations and assumptions. Our
experiments confirmed that guards are a very common
operation in traces. At the same time guards are associated with a high
overhead, because for all compiled guards information needs to be
stored to restore the execution state in case of a bailout. The measurements
showed that the compression techniques used in PyPy effectively reduce the
overhead of guards, but it still produces a significant overhead. The results
also showed that guard failure is a local event: there are few
guards that fail at all, and even fewer that fail very often.
These numbers validate the design decision of reducing the overhead of
successful guard checks as much as possible while paying a higher price in the
case of bailout due to having to decode a compressed state representation.
The compressed state representation reduces the memory footprint of rarely
used data.

@@

\section*{Acknowledgements}
We would like to thank David Edelsohn and Stephan Zalewski for their helpful
feedback and valuable comments while writing this paper.

>>>>

All measurements presented in this section do not take garbage collection of resume data and machine code into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and resume data that is generated.

@@

Figure~\ref{fig:benchmarks} extends Figure~\ref{fig:guard_percent} and summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The Figure also shows the overall optimization rate for
operations, which is between 69.4\% and 83.89\%, of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations. This indicates that the optimizer can remove
most of the guards, but after the optimization pass these still account for
15.2\% to 20.2\% of the operations being compiled and later executed.
The frequency of guard operations makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
quickly.

@@

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the compiled traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
enough that a bridge is compiled for them, only 1.2\% to 3.6\% of all guards
fail often enough that a bridge is compiled. Also, of all failing guards a few fail extremely often
and most fail rarely. The results emphasize that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

@@

SPUR~\cite{bebenita_spur:_2010} is a tracing JIT compiler
for a CIL virtual machine.
It handles guards by always generating code for every one of them
that transfers control back to the unoptimized code.
Since the transfer code needs to reconstruct the stack frames
of the unoptimized code,
the transfer code is quite large.

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a guard failure using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions plans to switch to compressed snapshots to further reduce
redundancy. The approach of not creating snapshots at all for every guard is
orthogonal to the resume data compression presented in this paper and could be
reused within RPython to improve the memory usage further.

Linking side exits to pieces of later compiled machine code was described first
in the context of Dynamo~\cite{Bala:2000wv} under the name of fragment linking.
Once a new hot trace is emitted into the fragment cache it is linked to the side
exit that led to the compilation of the fragment. Fragment Linking avoids the
performance penalty involved in leaving the compiled code. Fragment linking
also allows to remove compensation code associated to the linked fragments that
would have been required to restored the execution state on the side exit.

Gal et. al~\cite{Gal:2006} describe how in the HotpathVM, a JIT for a Java VM, they experimented
with having one generic compensation code block, like the RPython JIT, that
uses a register variable mapping to restore the interpreter state. Later this
was replaced by generating compensation code for each guard which produced a
lower overhead in their benchmarks. HotpathVM also records secondary traces
starting from failing guards that are connected directly to the original trace.
Secondary traces are compiled by first restoring the register allocator state to
the state at the side exit. The information is retrieved from a mapping stored
in the guard that maps machine level registers and stack to Java level stack
and variables.

@@

\section{Conclusion}
\label{sec:Conclusion}
In this paper we have concentrated on guards, an operation found in
tracing just-in-time compilers and used to denote points of possible control
flow divergence in recorded traces.
Based on the observation that guards are a frequent operation in traces and
that they do not fail often, we described how they have been implemented in the
high- and low-level components of RPython's tracing JIT compiler.

Additionally we have presented experimental data collected using the standard PyPy
benchmark set to evaluate previous observations and assumptions about guards. Our
experiments confirmed that guards are a very common
operation in traces. At the same time guards are associated with a high
overhead, because for all compiled guards information needs to be
stored to restore the execution state in case of a bailout. The measurements
showed that the compression techniques used in PyPy effectively reduce the
overhead of guards, but they still produce a significant overhead. The results
also showed that guard failure is a local event: there are few
guards that fail at all, and even fewer that fail very often.
These numbers validate the design decision of reducing the overhead of
successful guard checks as much as possible while paying a higher price in the
case of bailout due to having to decode a compressed state representation.
The compressed state representation reduces the memory footprint of rarely
used data.

@@

\section*{Acknowledgements}
We would like to thank David Edelsohn, Samuele Pedroni and Stephan Zalewski for their helpful
feedback and valuable comments while writing this paper.
